---
title: "Homework 2"
author: "Lupe Antonio"
date: "10/2/2023"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(readxl)
```


## Problem 1

```{r, message=FALSE}
#loading pols-month data
pols_month <- read_csv("data_hw2/fivethirtyeight_datasets/pols-month.csv")
```

```{r}
#cleaning the data
pols_month <- janitor::clean_names(pols_month) %>%
  #breaking up variable mon into integer variables: year, month, & day
  separate(mon, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december"),
    #creating president variable
    president = case_when(
      prez_dem == 1 & prez_gop == 0 ~ "democratic",
      prez_dem == 0 & prez_gop == 1 ~ "republican")) %>%
  #removing prez_dem, prez_gop, and day variables
  select(-prez_dem, -prez_gop, -day)
```


```{r, message=FALSE}
#loading snp.csv data
snp <- read_csv("data_hw2/fivethirtyeight_datasets/snp.csv") %>%
  #cleaning data
  janitor::clean_names() %>%
  #fixing the dates
  mutate(date = lubridate::mdy(date)) %>%
  #breaking up date variable into integer variables: year, month, day
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december")) %>%
  #removing day variable
  select(-day)
```


```{r, message=FALSE}
#loading unemployment.csv data
unemployment <- read_csv("data_hw2/fivethirtyeight_datasets/unemployment.csv") %>%
  #cleaning the data
  janitor::clean_names() %>%
  #going from "wide" to "long" format
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemp_perctg") %>%
  #fixing month names
  mutate(
    month = case_match(
      month,
      "jan" ~ "january",
      "feb" ~ "february",
      "mar" ~ "march",
      "apr" ~ "april",
      "may" ~ "may",
      "jun" ~ "june",
      "jul" ~ "july",
      "aug" ~ "august",
      "sep" ~ "september",
      "oct" ~ "october",
      "nov" ~ "november",
      "dec" ~ "december"
    ))
```


```{r}
#merging snp into pols
final_data <- left_join(pols_month, snp, by = c("year", "month"))

final_data <- left_join(final_data, unemployment, by = c("year", "month"))
```

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).






## Problem 2


Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?


```{r, message=FALSE}
#loading Mr. Trash Wheel data
mr_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #removing x15 & x16 columns
  select(-x15, -x16) %>%
  #creating new homes_powered variable
  mutate(
    homes_powered_2 = (weight_tons * 500)/30)
```


```{r}
#loading Professor Trash Wheel data
professor_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Professor Trash Wheel") %>%
  #cleaning data
  janitor::clean_names()
```


```{r}
#loading Gwynnda data
gwynnda_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Gwynnda Trash Wheel") %>%
  #cleaning data
  janitor::clean_names()
```

