---
title: "Homework 2"
author: "Lupe Antonio"
date: "10/2/2023"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```


## Problem 1

```{r, message=FALSE}
#loading pols-month data
pols_month <- read_csv("data_hw2/fivethirtyeight_datasets/pols-month.csv")
```

```{r}
#cleaning the data
pols_month <- janitor::clean_names(pols_month) %>%
  #breaking up variable mon into integer variables: year, month, & day
  separate(mon, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december"),
    #creating president variable
    president = case_when(
      prez_dem == 1 & prez_gop == 0 ~ "democratic",
      prez_dem == 0 & prez_gop == 1 ~ "republican")) %>%
  #removing prez_dem, prez_gop, and day variables
  select(-prez_dem, -prez_gop, -day)
```


```{r, message=FALSE}
#loading snp.csv data
snp <- read_csv("data_hw2/fivethirtyeight_datasets/snp.csv") %>%
  #cleaning data
  janitor::clean_names() %>%
  #fixing the dates
  mutate(date = lubridate::mdy(date)) %>%
  #breaking up date variable into integer variables: year, month, day
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december")) %>%
  #removing day variable
  select(-day)
```


```{r, message=FALSE}
#loading unemployment.csv data
unemployment <- read_csv("data_hw2/fivethirtyeight_datasets/unemployment.csv") %>%
  #cleaning the data
  janitor::clean_names() %>%
  #going from "wide" to "long" format
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemp_perctg") %>%
  #fixing month names
  mutate(
    month = case_match(
      month,
      "jan" ~ "january",
      "feb" ~ "february",
      "mar" ~ "march",
      "apr" ~ "april",
      "may" ~ "may",
      "jun" ~ "june",
      "jul" ~ "july",
      "aug" ~ "august",
      "sep" ~ "september",
      "oct" ~ "october",
      "nov" ~ "november",
      "dec" ~ "december"
    ))
```


```{r}
#merging snp into pols
final_data <- left_join(pols_month, snp, by = c("year", "month"))

final_data <- left_join(final_data, unemployment, by = c("year", "month"))
```

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).






## Problem 2


Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda, and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?


```{r, message=FALSE}
#loading Mr. Trash Wheel data
mr_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #removing x15 & x16 columns
  select(-x15, -x16) %>%
  #creating new homes_powered variable
  mutate(
    homes_powered = (weight_tons * 500)/30) %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Mr. Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Professor Trash Wheel data
professor_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Professor Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Professor Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Gwynnda data
gwynnda_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Gwynnda Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Gwynnda Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#combining datasets
trash_wheels_tidy <- bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda_trash_wheel)
```


Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?




## Problem 3

Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.


```{r, message=FALSE}
#loading the baseline dataset
baseline_demos <- read_csv("data_hw2/data_mci/MCI_baseline.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  #renaming the column names
  rename(
    'id' = 'x1',
    'baseline_age' = 'age_at_the_study_baseline',
    'sex' = 'x1_male_0_female',
    'education_years' = 'years_of_education',
    'apoe4_carrier' = 'x1_apoe4_carrier_0_apoe4_non_carrier',
    'onset_age' = 'age_at_the_onset_of_mci_missing_if_a_subject_remains_mci_free_during_the_follow_up_period')
```


```{r, message=FALSE}
#continuation of tidying data
baseline_demos <- baseline_demos %>%
  mutate(
    #changing sex values
    sex = case_when(
      sex == 0 ~ "female",
      sex == 1 ~ "male"),
    #changing APOE4 carrier values
    apoe4_carrier = case_when(
      apoe4_carrier == 0 ~ "non-carrier",
      apoe4_carrier == 1 ~ "carrier"),
    #adding NA's in blank spaces
    onset_age = case_when(
      onset_age == "." ~ NA,
      TRUE ~ as.numeric(onset_age)
    )) %>%
  #filtering no MCI
  filter(onset_age > baseline_age | is.na(onset_age))
```



```{r, message=FAlSE}
#loading other datasets
amyloid_dataset <- read_csv("data_hw2/data_mci/mci_amyloid.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  
```


