---
title: "Homework 2"
author: "Lupe Antonio"
date: "10/2/2023"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```


## Problem 1

```{r, message=FALSE}
#loading pols-month data
pols_month <- read_csv("data_hw2/fivethirtyeight_datasets/pols-month.csv")
```

```{r}
#cleaning the data
pols_month <- janitor::clean_names(pols_month) %>%
  #breaking up variable mon into integer variables: year, month, & day
  separate(mon, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december"),
    #creating president variable
    president = case_when(
      prez_dem == 1 & prez_gop == 0 ~ "democratic",
      prez_dem == 0 & prez_gop == 1 ~ "republican")) %>%
  #removing prez_dem, prez_gop, and day variables
  select(-prez_dem, -prez_gop, -day)
```


```{r, message=FALSE}
#loading snp.csv data
snp <- read_csv("data_hw2/fivethirtyeight_datasets/snp.csv") %>%
  #cleaning data
  janitor::clean_names() %>%
  #fixing the dates
  mutate(date = lubridate::mdy(date)) %>%
  #breaking up date variable into integer variables: year, month, day
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #fixing the year problem
    year = ifelse(year > 2023, year - 100, year),
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december")) %>%
  #removing day variable
  select(-day)
```
 

```{r, message=FALSE}
#loading unemployment.csv data
unemployment <- read_csv("data_hw2/fivethirtyeight_datasets/unemployment.csv") %>%
  #cleaning the data
  janitor::clean_names() %>%
  #going from "wide" to "long" format
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemp_perctg") %>%
  #fixing month names
  mutate(
    month = case_match(
      month,
      "jan" ~ "january",
      "feb" ~ "february",
      "mar" ~ "march",
      "apr" ~ "april",
      "may" ~ "may",
      "jun" ~ "june",
      "jul" ~ "july",
      "aug" ~ "august",
      "sep" ~ "september",
      "oct" ~ "october",
      "nov" ~ "november",
      "dec" ~ "december"
    ))
```


```{r}
#merging snp into pols
final_data <- left_join(pols_month, snp)

final_data <- left_join(final_data, unemployment)
```


The `pols_month` data has a total of `r nrow(pols_month)` observations and `r ncol(pols_month)` variables. This dataset contains information regarding the party affiliations of governors and senators, and the current president (at that time) between `r pols_month %>% pull(year) %>% min()` to `r pols_month %>% pull(year) %>% max()`.
The `snp` data has a total of `r nrow(snp)` observations and `r ncol(snp)` variables. It contains information from the years `r snp %>% pull(year) %>% min()` to `r snp %>% pull(year) %>% max()`. 
Additionally, the `unemployment` dataset has a total of `r nrow(unemployment)` observations and `r ncol(unemployment)` variables. It contains information from the years `r unemployment %>% pull(year) %>% min()` to `r unemployment %>% pull(year) %>% max()`. 





## Problem 2

```{r, message=FALSE}
#loading Mr. Trash Wheel data
mr_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #removing x15 & x16 columns
  select(-x15, -x16) %>%
  #creating new homes_powered variable
  mutate(
    homes_powered = (weight_tons * 500)/30) %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Mr. Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Professor Trash Wheel data
professor_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Professor Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Professor Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Gwynnda data
gwynnda_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Gwynnda Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Gwynnda Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#combining datasets
trash_wheels_tidy <- bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda_trash_wheel)
```



The final tidy dataset contains information from `mr_trash_wheel`, `professor_trash_wheel`, and `gwynnda_trash_wheel`. It has a total of `r nrow(trash_wheels_tidy)` observations and `r ncol(trash_wheels_tidy)` variables. This dataset contains information regarding the amount of trash per different trash type between `r trash_wheels_tidy %>% pull(year) %>% min()` and `r trash_wheels_tidy %>% pull(year) %>% max()` for each type of trash wheel. 

From the available data, the total weight (in tons) of trash collected by Professor Trash Wheel was `r professor_trash_wheel %>% sum(weight_tons)`. The total weight (in tons) of trash collected by Mr. Trash Wheel was `r mr_trash_wheel %>% sum(weight_tons)`. Simiarly, the total weight (in tons) of trash collected by Gwynnda Trash Wheel was `r gwynnda_trash_wheel %>% sum(weight_tons)`.
Additionally, the total number of cigarette butts collected by Gwynnda in July of 2021 was `r gwynnda_trash_wheel %>% filter(month == "July" & year == "2021") %>% sum(cigarette_butts)`.  



## Problem 3


```{r, message=FALSE}
#loading the baseline dataset
baseline_demos <- read_csv("data_hw2/data_mci/MCI_baseline.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  #renaming the column names
  rename(
    'baseline_age' = 'current_age',
    'education_years' = 'education',
    'apoe4_carrier' = 'apoe4',
    'onset_age' = 'age_at_onset')
```


```{r, message=FALSE}
#continuation of tidying data
baseline_demos <- baseline_demos %>%
  mutate(
    #changing sex values
    sex = case_when(
      sex == 0 ~ "female",
      sex == 1 ~ "male"),
    #changing APOE4 carrier values
    apoe4_carrier = case_when(
      apoe4_carrier == 0 ~ "non-carrier",
      apoe4_carrier == 1 ~ "carrier"),
    #adding NA's in blank spaces
    onset_age = case_when(
      onset_age == "." ~ NA,
      TRUE ~ as.numeric(onset_age)
    )) %>%
  #filtering no MCI
  filter(onset_age > baseline_age | is.na(onset_age))
```
Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers? ^^





```{r, message=FAlSE}
#loading other datasets
amyloid_dataset <- read_csv("data_hw2/data_mci/mci_amyloid.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  #renaming colname
  rename('id' = 'study_id') %>%
  #transforming dataset
  pivot_longer(baseline:time_8,
               names_to = 'time_elapsed',
               values_to = 'biomarkers_ratio')
```

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset. ^




```{r}
#are in baseline
anti_join(baseline_demos, amyloid_dataset, by = 'id')
```

```{r}
#aren't in baseline
anti_join(amyloid_dataset, baseline_demos, by = 'id')
```

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. ^^ (both)




```{r}
#dataset with common observations
common_amyloid_dataset <- inner_join(baseline_demos, amyloid_dataset, by = "id")

#export csv
write_csv(common_amyloid_dataset,"data_hw2/common_amyloid_dataset.csv")
```

briefly describe the resulting dataset^