---
title: "Homework 2"
author: "Lupe Antonio"
date: "10/2/2023"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```


## Problem 1

```{r, message=FALSE}
#loading pols-month data
pols_month <- read_csv("data_hw2/fivethirtyeight_datasets/pols-month.csv")
```

```{r}
#cleaning the data
pols_month <- janitor::clean_names(pols_month) %>%
  #breaking up variable mon into integer variables: year, month, & day
  separate(mon, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december"),
    #creating president variable
    president = case_when(
      prez_dem == 1 & prez_gop == 0 ~ "democratic",
      prez_dem == 0 & prez_gop == 1 ~ "republican")) %>%
  #removing prez_dem, prez_gop, and day variables
  select(-prez_dem, -prez_gop, -day)
```


```{r, message=FALSE}
#loading snp.csv data
snp <- read_csv("data_hw2/fivethirtyeight_datasets/snp.csv") %>%
  #cleaning data
  janitor::clean_names() %>%
  #fixing the dates
  mutate(date = lubridate::mdy(date)) %>%
  #breaking up date variable into integer variables: year, month, day
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    #fixing the year problem
    year = ifelse(year > 2023, year - 100, year),
    #replacing month num with month name
    month = case_match(
      month,
      1 ~ "january",
      2 ~ "february",
      3 ~ "march", 
      4 ~ "april",
      5 ~ "may",
      6 ~ "june",
      7 ~ "july",
      8 ~ "august",
      9 ~ "september",
      10 ~ "october",
      11 ~ "november",
      12 ~ "december")) %>%
  #removing day variable
  select(-day)
```
 

```{r, message=FALSE}
#loading unemployment.csv data
unemployment <- read_csv("data_hw2/fivethirtyeight_datasets/unemployment.csv") %>%
  #cleaning the data
  janitor::clean_names() %>%
  #going from "wide" to "long" format
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemp_perctg") %>%
  #fixing month names
  mutate(
    month = case_match(
      month,
      "jan" ~ "january",
      "feb" ~ "february",
      "mar" ~ "march",
      "apr" ~ "april",
      "may" ~ "may",
      "jun" ~ "june",
      "jul" ~ "july",
      "aug" ~ "august",
      "sep" ~ "september",
      "oct" ~ "october",
      "nov" ~ "november",
      "dec" ~ "december"
    ))
```


```{r}
#merging snp into pols
final_data <- left_join(pols_month, snp)

final_data <- left_join(final_data, unemployment)
```


The `pols_month` data has a total of `r nrow(pols_month)` observations and `r ncol(pols_month)` variables. This dataset contains information regarding the party affiliations of governors and senators, and the current president (at that time) between `r pols_month %>% pull(year) %>% min()` to `r pols_month %>% pull(year) %>% max()`.
The `snp` data has a total of `r nrow(snp)` observations and `r ncol(snp)` variables. It contains information from the years `r snp %>% pull(year) %>% min()` to `r snp %>% pull(year) %>% max()`. 
Additionally, the `unemployment` dataset has a total of `r nrow(unemployment)` observations and `r ncol(unemployment)` variables. It contains information from the years `r unemployment %>% pull(year) %>% min()` to `r unemployment %>% pull(year) %>% max()`. 





## Problem 2

```{r, message=FALSE}
#loading Mr. Trash Wheel data
mr_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #removing x15 & x16 columns
  select(-x15, -x16) %>%
  #creating new homes_powered variable
  mutate(
    homes_powered = (weight_tons * 500)/30) %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Mr. Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Professor Trash Wheel data
professor_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Professor Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Professor Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#loading Gwynnda data
gwynnda_trash_wheel <- read_excel("data_hw2/Trash_Wheel_Collection_Data.xlsx", sheet = "Gwynnda Trash Wheel") %>%
  #cleaning data
  janitor::clean_names() %>%
  #adding trash wheel variable
  mutate(trash_wheel = "Gwynnda Trash Wheel") %>%
  #filter NA's in dumpster
  filter(!is.na(dumpster)) %>%
  #making year numeric
  mutate(year = as.numeric(year))
```


```{r}
#combining datasets
trash_wheels_tidy <- bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda_trash_wheel)
```


The final tidy dataset contains information from `mr_trash_wheel`, `professor_trash_wheel`, and `gwynnda_trash_wheel`. It has a total of `r nrow(trash_wheels_tidy)` observations and `r ncol(trash_wheels_tidy)` variables. This dataset contains information regarding the amount of trash per different trash type between `r trash_wheels_tidy %>% pull(year) %>% min()` and `r trash_wheels_tidy %>% pull(year) %>% max()` for each type of trash wheel. 

From the available data, the total weight (in tons) of trash collected by Professor Trash Wheel was `r professor_trash_wheel %>% pull(weight_tons) %>% sum()`. The total weight (in tons) of trash collected by Mr. Trash Wheel was `r mr_trash_wheel %>% pull(weight_tons) %>% sum()`. Simiarly, the total weight (in tons) of trash collected by Gwynnda Trash Wheel was `r gwynnda_trash_wheel %>% pull(weight_tons) %>% sum()`.
Additionally, the total number of cigarette butts collected by Gwynnda in July of 2021 was `r gwynnda_trash_wheel %>% filter(month == "July" & year == "2021") %>% pull(cigarette_butts) %>% sum()`.  



## Problem 3


```{r, message=FALSE}
#loading the baseline dataset
baseline_demos <- read_csv("data_hw2/data_mci/MCI_baseline.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  #renaming the column names
  rename(
    'baseline_age' = 'current_age',
    'education_years' = 'education',
    'apoe4_carrier' = 'apoe4',
    'onset_age' = 'age_at_onset')
```


```{r, message=FALSE}
#continuation of tidying data
baseline_demos <- baseline_demos%>%
  mutate(
    #changing sex values
    sex = case_when(
      sex == 0 ~ "female",
      sex == 1 ~ "male"),
    #changing APOE4 carrier values
    apoe4_carrier = case_when(
      apoe4_carrier == 0 ~ "non-carrier",
      apoe4_carrier == 1 ~ "carrier"),
    #adding NA's in blank spaces
    onset_age = case_when(
      onset_age == "." ~ NA,
      TRUE ~ as.numeric(onset_age)
    )) %>%
  #filtering no MCI
  filter(onset_age > baseline_age | is.na(onset_age))
```


The `baseline_demos` dataset was imported and tidyied up by renaming column names when necessary, mutating the `sex` variable values into female or male, and similarly the `apoe4_carrier` into carrier or non-carrier. Additionally, NA's were needed to `onset_age` variable, and we filtered for participants without MCI at baseline. 
The resulting dataset contains a total of `r nrow(baseline_demos)` observations and `r ncol(baseline_demos)` variables. There was a total of 483 participants recruited and of those `r nrow(baseline_demos)` developed MCI. The average baseline age was `r baseline_demos %>% pull(baseline_age) %>% mean()`. The proportion of women in the study that were APOE4 carriers were `r baseline_demos %>% filter(sex == 'female' & apoe4_carrier == "carrier") %>% pull(id) %>% sum()` / `r nrow(baseline_demos)`.  





```{r, message FALSE}
#loading other datasets
amyloid_dataset <- read_csv("data_hw2/data_mci/mci_amyloid.csv", skip = 1) %>%
  #cleaning dataset
  janitor::clean_names() %>%
  #renaming colname
  rename('id' = 'study_id') %>%
  #transforming dataset
  pivot_longer(baseline:time_8,
               names_to = 'time_elapsed',
               values_to = 'biomarkers_ratio')
```


The `amyloid_dataset` dataset was imported and tidyied up by renaming column names when necessary, and transforming the dataset from 'wide' format to 'long' format. 
The resulting dataset contains a total of `r nrow(amyloid_dataset)` observations and `r ncol(amyloid_dataset)` variables.


```{r}
#observations in the baseline_demos dataset, but not in amyloid_dataset
anti_join(baseline_demos, amyloid_dataset, by = 'id')
```

Of the participants that were in the `baseline_demos` dataset, most were non-carriers of APOE4, and most of their onset age is missing. 



```{r}
#observations in amyloid_dataset, but not in baseline_demos dataset
anti_join(amyloid_dataset, baseline_demos, by = 'id')
```

Based on these findings, we can see that the participants that were not in the `baseline_demos`has simialr biomarker ratios.





```{r}
#dataset with common observations
common_amyloid_dataset <- inner_join(baseline_demos, amyloid_dataset, by = "id")

#export csv
write_csv(common_amyloid_dataset,"data_hw2/common_amyloid_dataset.csv")
```

After combining both datasets, the `common_amyloid_dataset` is produced. This dataset contains at total of `r nrow(common_amyloid_dataset)` observations and `r ncol(common_amyloid_dataset)` variables. This dataset contains information of participants' biomarkers ratio obtained at different times/stages in the study, and shows whether the participant is a APOE4 carrier or not, the participants' sex, ages at baseline and onset, and their years of education. 
\